### 2015-05-08 whoop.tex
First draft of the simple programming language with informal semantics.

In short, symbolic pairwise lockset analysis works by considering, for a given device driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race, accounting for the effects of additional entry points by treating the driver shared state abstractly so that when an entry point reads from the shared state, a nondeterministic value is returned.  Symbolic verification of an entry point pair works by instrumenting each entry point with additional state to record locksets (see Section~\ref{bg:lockset}), and attempting to verify a sequential program that executes the instrumented entry points in sequence, and subsequently asserts that the lockset for each shared variable is non-empty.  Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.

@@

\paragraph{A simple concurrent programming model}

We consider a concurrent programming model in which an unbounded number of threads execute a set of pre-defined \emph{thread templates}.  In the context of device drivers, a thread template corresponds to a driver entry point.  At any given point of execution a certain number of threads are active, each thread executing a particular template.  As in the case of drivers, where multiple distinct entry points as well as multiple instances of the same entry point may run concurrently, two distinct threads in our model may execute according to the same template or according to different templates.  Further threads may be launched at any point during execution, just as in the context of device drivers where the OS may invoke additional driver entry points.  For ease of presentation only, our model does not feature aggregate data types, pointers or dynamic memory allocation.  These \emph{are} handled by our implementation, and in Section~\ref{sec:whoopimplementation} we discuss interesting practical issues arising from the handling of a full-blown language.

A \emph{concurrent program} is described by a finite set of \emph{shared} variables, $\sharedvars$, a finite set of mutexes, $\mutexes$, and a finite set of thread templates, $T_1,\dots,T_n$ (for some $n > 0$).  A thread template, $T$, is described by a finite set of procedures, $\procedures{T}$, and a finite set of thread-private variables, $\variables{T}$.  A designated procedure, procedure $\mainforthread{T} \in \procedures{T}$, denotes the starting point for execution of a template by a thread.  A procedure $P \in \procedures{T}$ is represented by a control flow graph of basic blocks, where each block contains a sequence of statements.  A basic block either has a single successor or a pair of successors.  In the latter case, an \emph{exit condition} over thread-private variables determines, at runtime, the successor to which control should flow on block exit.

The forms of statements are shown in Figure~\ref{fig:statements}, and include designated statements for reading from and writing to shared variables.  In particular, shared variables may not appear in arbitrary expressions.  This restriction simplifies our presentation of lockset instrumentation, and a program that does not satisfy this restriction can be trivially pre-processed into one that does through the introduction of additional private temporaries to record values read from the shared state.  We do not specify the form of expressions, nor the types of variables, assuming a standard set of data types and operations.

@@

\ADComment{Say something clear, but intuitive rather than precise, about the semantics.}

\paragraph{Data races}

\ADComment{Say what a data race is, and address the weak memory issue.}

\paragraph{Lockset instrumentation}

>>>>

Symbolic pairwise lockset analysis works by considering, for a given device driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race, accounting for the effects of additional entry points by treating the driver shared state abstractly so that when an entry point reads from the shared state, a nondeterministic value is returned.  Symbolic verification of an entry point pair works by instrumenting each entry point with additional state to record locksets (see Section~\ref{bg:lockset}), and attempting to verify a sequential program that executes the instrumented entry points in sequence, and subsequently asserts that the lockset for each shared variable is non-empty.  Verification of the resulting sequential program can be undertaken using any sound method; in practice we employ the Boogie verification engine~\cite{barnett2006boogie}, which requires procedure specifications and loop invariants to be generated, after which verification conditions are generated and discharged to an automated theorem prover.

@@

\medskip\noindent\textbf{Concurrent programming model }
%
We consider a simple concurrent programming model in which an
unbounded number of threads execute a set of pre-defined \emph{thread
  templates}.  At any given point of execution a certain number of
threads are active, each thread executing a particular template.  In
the context of device drivers, a thread template corresponds to a
driver entry point, and multiple instances of the same thread template
may execute concurrently, just as multiple invocations of a single
driver entry point may be concurrent.  Further threads may be launched
at any point during execution; in the context of device drivers this
corresponds to the OS invoking additional driver entry points.  For
ease of presentation only, our model does not feature aggregate data
types, pointers or dynamic memory allocation.  These \emph{are}
handled by our implementation, and in
Section~\ref{sec:whoopimplementation} we discuss interesting practical
issues arising from the handling of a full-blown language.

A \emph{concurrent program} is described by a finite set of
\emph{shared} variables, $\sharedvars$, a finite set of mutexes,
$\mutexes$, and a finite set of \emph{thread templates}.  A thread
template, $T$, consists of a finite set of procedures,
$\procedures{T}$, and a finite set of private variables,
$\variables{T}$.  A designated procedure, procedure $\mainforthread{T}
\in \procedures{T}$, denotes the starting point for execution of a
template by a thread.  A procedure $P \in \procedures{T}$ is
represented by a control flow graph of basic blocks, where each block
contains a sequence of statements.  A basic block either has a single
successor or a pair of successors.  In the latter case, an \emph{exit
  condition} over thread-private variables determines, at runtime, the
successor to which control should flow on block exit.

The forms of statements are shown in Figure~\ref{fig:statements}, and
include designated statements for reading from and writing to shared
variables.  In particular, shared variables may not appear in
arbitrary expressions.  This restriction simplifies our presentation
of lockset instrumentation, below, and a program that does not satisfy
this restriction can be trivially pre-processed into one that does
through the introduction of additional private temporaries to record
values read from the shared state.  We do not specify the form of
expressions, nor the types of variables, assuming a standard set of
data types and operations.

@@

\medskip\noindent\textbf{Semantics} Let $\ids$ be an infinite set from
which dynamic thread ids will be drawn.  The state of a running
concurrent program consists of: a valuation of the shared variables,
$\sharedvars$; a mapping that associates each mutex in $\mutexes$ with
an id from $\ids$, recording which thread currently holds the mutex,
or with a special value $\bot \notin \ids$ to indicate that the
mutex is not held by any thread; a list of \emph{threads}.  Each
thread consists of an id, drawn from $\ids$, a thread template, $T$,
an index indicating the next statement of the template $T$ to be
executed by the thread, and a valuation of the thread private
variables, $\variables{T}$.  If multiple threads are instances of the
same template $T$, then each thread carries a \emph{separate}
valuation of the private variables for this template.

At the start of execution the valuation of shared variables is
arbitrary, no mutexes are held (i.e.\ each mutex maps to $\bot$),
and the list of threads is empty.

At any point of execution, a new thread may be added to the list of
thread states.  This involves selecting an id $i \in \ids$ that has
not been previously used during program execution and a thread
template $T$, setting the point of execution for the new thread to be
the first statement of $\mainforthread{T}$, and choosing an arbitrary
valuation for the private variables $\variables{T}$.

We consider a standard interleaving model of concurrency: at any
execution point, a thread may execute its current statement, unless
that statement has the form $\mutexlock(m)$ and mutex $m$ is already held
by some thread.  Executing a statement causes the state of the thread,
and the shared state, to be updated in a standard manner.  For
example, if a thread following template $T$ executes $s = e$, where $s
\in \sharedvars$ and $e$ is an expression over $\variables{T}$, the
shared variable valuation is updated so that $s$ has the value
determined by evaluating $e$ in the context of the thread's private
variable valuation.  The thread's next statement is updated according
to the control flow graph of the current procedure.

A thread terminates if it reaches the end of $\mainforthread{T}$; in this
case the thread is removed from the list of threads.  Because we are
interested in the analysis of device drivers, which are
\emph{reactive} concurrent programs, we do not consider the notion of
global program termination.

\medskip\noindent\textbf{Data races } \ADComment{We now need to define
what a data race is.  This could be done using happens-before.  I am undecided as to whether
we should refer back to the happens-before material in the background section, or whether
we should move that material here, so that we explain what happens-before means, and then
explain what a data races is.  My preference is for the latter: doing it all here.}

\medskip\noindent\textbf{Lockset instrumentation} \ADComment{This is as far as I have gotten with the writing.  I now want to explain formally and somewhat abstractly how our lockset instrumentation works.}
